Leadership across cultures and contexts: an exploratory analysis

### 2020-05-14 leadership_across_cultures_contexts.Rmd
minor change in conclusion.

Leaders, possibly due to their special capacities, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

## The five-fold way

In our view, the dominance-prestige and "dual-model" models of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "managers". In short, these models have conflated "benefits" with knowledge and expertise, and "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006].

@@

## Possible ethnographic or publication biases

asf

@@

Our results also suggest both broad continuities and contrasts between leadership among the nonindustrial, non-WEIRD populations sampled here and the WEIRD samples common in managerial studies. Our organize and strategize features of leader functions reflect the @wofford_factor_1970 *order and group achievement* and *personal interaction* factors, respectively. Also, Woffords *personal enhancement* and *dynamic and achievement* factors are similar to leadership dimensions captured by our competencies feature. Our results, however, emphasize the prosociality of leaders in non-WEIRD settings over an emphasis on self-preservation, relative to WEIRD leaders (e.g., Woffords *security and maintenance* factor). We also identified conformity to social norms as a feature of human leadership across cultures, similar to predictions and findings drawing on Culturally Endorsed Implicit Leadership Theory by managerial scholars working with WEIRD populations [@behrendt_integrative_2017]. Our results on the functions of leaders across cultures and contexts also corresponds to the *task-oriented* vs. *relation-oriented* framework of managerial behaviors (i.e., our Management vs. Prosociality features of leader functions), however, our results do not support the strong emphasis on charismatic leadership found by the GLOBE CEO study [@house2013strategic] (although charisma was documented as a quality of leaders in over 20% of cultures and oratory skills in over 50%). See Figure \@ref(fig:cluster).

These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographic record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models behavior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely fails to capture much of the ethnographic reality of human leadership dynamics.

>>>>

Leaders, possibly due to their special competencies, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 [p. 185] describes the nature of conflict resolution processes in the absence of strong authority:

@@

## The five-fold way OR Leadership markets

In our view, the dominance-prestige and "dual-model" theories of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "managers". In short, these models have conflated "benefits" with knowledge and expertise, and "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006].

@@

@@

<!-- CONNECT TO WERID -->
Our results also suggest both broad continuities and contrasts between leadership among the nonindustrial, non-WEIRD populations sampled here and the WEIRD samples common in managerial studies. Our organize and strategize features of leader functions reflect the @wofford_factor_1970 *order and group achievement* and *personal interaction* factors, respectively. Also, Woffords *personal enhancement* and *dynamic and achievement* factors are similar to leadership dimensions captured by our competencies feature. Our results, however, emphasize the prosociality of leaders in non-WEIRD settings over an emphasis on self-preservation, relative to WEIRD leaders (e.g., Woffords *security and maintenance* factor). We also identified conformity to social norms as a feature of human leadership across cultures, similar to predictions and findings drawing on Culturally Endorsed Implicit Leadership Theory by managerial scholars working with WEIRD populations [@behrendt_integrative_2017]. Our results on the functions of leaders across cultures and contexts also corresponds to the *task-oriented* vs. *relation-oriented* framework of managerial behaviors (i.e., our Management vs. Prosociality features of leader functions), however, our results do not support the strong emphasis on charismatic leadership found by the GLOBE CEO study [@house2013strategic] (although charisma was documented as a quality of leaders in over 20% of cultures and oratory skills in over 50%). See Figure \@ref(fig:cluster).

<!-- SUMMARY AND SUGGEST LEADERS BE PART OF OTHER GROUP RESEARCH -->
These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographic record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models behavior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

<!-- CLOSING STATEMENT -->
Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely fails to capture much of the ethnographic reality of human leadership dynamics.
